\Huge Conferences in an Era of Expensive Carbon

### 2019-06-10 mike.tex
Reworked.

% \begin{multicols}{3}

ACM's vibrant conference culture has a dark side: it contributes
significantly to climate change.

% \begin{figure}[t]
% \centering
% \includegraphics[width=4in]{wmo_stripes.png}
% \caption{Variation in annual global temperature, 1850 to 2018 (range from
%   blue to red: $1.35^\circ$C) \cite{WarmingStripes}}
% \label{fig:warming}
% \end{figure}

% \begin{figure}[t]
% \centering
% \includegraphics[width=2in]{IPCC-cover.png}
% \caption{2018 IPCC report \cite{IPCC18}.  Key finding: Emissions must
%   decline by at least 40 percent by 2030 and reach net zero by 2050, if we
%   are to hold warming to 1.5 degrees.  There is ``no documented historic
%   precedent'' for the transformation of the world economy needed to achieve
%   this.  }
% \label{fig:IPCC-cover}
% \end{figure}

@@

Indeed, scientific societies arguably have an especially important role to
play, since one of their main activities is organizing international
conferences, which produce significant greenhouse gas emissions,
particularly from air travel. \mwh{The previous is the wrong
  transition, since again it seems to suggest that travel is a big
  deal in terms of the overall problem. The sentence also fails to
  convey the other reason scientific societies should be involved: As
  role models and advocates for general change.}  A single round-trip flight from
Philadelphia to Paris emits the equivalent of about 1.7 tons of carbon
dioxide (\COtwoE) per passenger~\cite{carboncalculator}\iflater\bcp{Trim
  citation?}\fi---a substantial fraction of the {\em total} yearly emissions
for an average resident of the US (16.5 tons) or Europe (7
tons)~\cite{emissions}.  \iflater\bcp{For comparison, ...  (put in some
  figures about the footprints of a few other things that people might think
  of reducing---driving, eating meat, upgrading their fridge...)}\fi{}
Moreover, these emissions have no near-term 
technological fix, since jet fuel is difficult to replace with
renewable energy sources~\cite{elec-air}. 
% \iflater\bcp{Trimmed this sentence: Reducing the \COtwoE{} footprint
% of conferences will require a reduction in the footprint of the attendees'
% air travel.}\fi

Conferences obviously also create significant value.  Science is a
fundamentally social process, and the conference system accelerates research
through high-bandwidth interaction, direct dissemination of results, network
building, and serendipitous cross-fertilization.

This leads to a puzzle: How do we weigh the environmental harm caused by
conference travel against the social benefits of the scientific progress
that comes from conferences?  It seems an extremely difficult question to
answer: the harm and the compensating benefits cannot be directly
compared---they are apples and oranges.

@@

Fortunately, there is a well-established mechanism for comparing apples and
oranges---indeed, for comparing almost anything!  Economists call it ``the
magic of markets.''  By mapping choices and preferences in disparate domains
onto a common scale (money), economic markets can be viewed as massively
distributed computations that optimize costs and benefits across vastly
different activities.\bcp{Give a simple, concrete example.}  For this
reason, many policy experts advocate addressing climate change through some
form of carbon pricing, which places a monetary cost on greenhouse gas
emissions, sending a clear market signal that changes aggregate
behavior.\bcp{See
  https://kleinmanenergy.upenn.edu/policy-digests/why-carbon-pricing-falls-short
  for lots of good phrasing and citations.}  Exactly how the price of carbon
should be set is a challenging policy question, but it has a clear goal: It
should be chosen so that the world arrives at zero emissions within the
timeframe that climate scientists tell us is safe.

To be effective in the long term, a carbon price needs to be uniform across
society---i.e., it should be imposed by governments.  Indeed, many
countries\bcp{e.g.?} and states (e.g., California) have already adopted
carbon pricing policies; more will follow.
%
Bottom line: We are entering an era in which carbon will be expensive.

Anticipating this era, what should we rationally do now to ensure that ACM,
and computer-science research more broadly, continue to thrive in a world
where air travel is more expensive?  Believing that it needs to come quickly
to avert disaster, what can we do to accelerate the process? \mwh{The
  question now seems to have changed. The paper started by saying ACM has an
  ``especially important role to play,'' but now seems to suggest it should
  be reacting to an inevitable/coming future. These questions probably need
  to go much earlier, and the earlier text reworked around them.}

\SECTION{Carbon Offsets}

In the short term, one possible response to both questions is to use {\em
  carbon offsets}, which allow organizations (and individuals) to
voluntarily impose a price on their own emissions~\cite{CarbonOFfsetReport}.
%
In a nutshell, the idea is for a conference to raise funds from
registration fees or corporate sponsors and donate them, via a carbon offset
{\em vendor}, to a {\em provider} who will use them to support an activity
(e.g., installing methane capture devices on landfill sites or buying
fuel-efficient stoves to replace open-fire cooking in poor communities) that
is {\em certified} (by yet another organization) to permanently avoid
emitting an amount of greenhouse gases equivalent to the emissions from
participants' travel to the conference.
%
Many organizations (including companies such as Google, Dell, Microsoft,
General Motors, Delta Airlines, Lyft, and Expedia, numerous universities,
and a number of academic societies) are already buying carbon offsets, and
some ACM conferences have already started to experiment with voluntary
carbon offset fees at registration time.

Carbon offsets do come with some significant concerns.
%
One superficial impediment is that most organizations do not currently have
accounting procedures in place to deal with them, and the piecemeal
political and bureaucratic work of designing and implementing a separate
policy for each organization can consume significant energy.
%
Another point to be careful of is that the long-term climate benefit of some
offsetting strategies (for example, paying people to plant trees) is
controversial.
%
A more fundamental problem is that it is not straightforward to decide how
much one should pay to offset a ton of carbon: current prices vary widely
around an average of around \$15 per ton, depending on the vendor, the
nature of the offset project, the region of the world where it will be
implemented, etc.  Moreover, in order for offsets (or any other form of a
carbon pricing) to fully motivate the needed changes in behavior, the price
may need to go much higher---the true ``social cost'' of a ton of carbon is
variously estimated at anywhere from \$40 to \$400 per
ton~\bcp{citation}---and setting it too low may encourage complacency.
%
However, while acknowledging these concerns, we believe that voluntary
carbon offsetting can be an effective near-term tool for raising awareness
and beginning to shift organizational behavior because it associates
emissions with a number, giving decision makers within the organization
something concrete to optimize.

\SECTION{Measure, Analyze, Optimize}

@@

Beyond imposing a price on our carbon footprint (or waiting for one
to be imposed from outside), a sane response to climate change should
clearly also include taking steps to reduce the size of this
footprint.
\mwh{Don't like previous sentence. It seems to have gotten away from
  the point of the paper, which is using price to change balance of
  apples to oranges. Analyzing data is useful for figuring out ACM's
  internal price, not tweaking around the edges, which is how it reads
  below.} For
conferences, there are many options that one might
contemplate~\cite{ClimateCommitteReport}, from easy 
but incremental changes like making program committee meetings virtual
instead of physical and enabling virtual participation via live-streaming,
all the way to complete re-imagining of conferences and their role in
science (holding conferences completely in virtual spaces, replacing
conferences with journals, etc.), with some intriguing hybrid forms in the
middle (regional conferences, mega-conferences, conferences split into
two locations on different continents, etc.).  
%
However, before we undertake large changes, we need to understand the
situation better.  For most of these ideas, it is hard to confidently
predict either the potential benefits (in terms of reduced emissions) or the
potential costs (to scientific exchange, career development of young
scientists, etc.).  Indeed, a preliminary analysis of registration data from
a number of SIGPLAN conferences paints a rather complicated picture.
%
\bcp{Put the discussion of Figure 1 (and an accompanying figure
showing distance traveled to several instances of the same conference) here;
this illustrates the difficulty and the need for careful data gathering and
analysis.} 
All parts of ACM should be doing this kind of analysis.  (The acm-climate
mailing list is one place to exchange ideas about this.)

\SECTION{Deeper Changes}

Ultimately, responding to the inevitability of expensive carbon may involve
rethinking the whole basis of how academic computer science is done.
\bcp{
\begin{itemize}
\item Add a bit of meat about how the current economic model may be
  unsustainable and reworking it will involve substantial organizational
  changes.  (There may be some good words that we can paraphrase from our
  second reviewer!)
\item monitoring and mitigating carbon costs should be an ACM-level activity
\end{itemize}
}

>>>>

% ACM's vibrant conference culture has a dark side: it contributes
% significantly to climate change.

@@

Indeed, scientific societies arguably have an especially important role to
play, since one of their main activities is organizing international
conferences, which produce significant greenhouse gas emissions,
particularly from air travel. A single round-trip flight from
Philadelphia to Paris emits the equivalent of about 1.7 tons of carbon
dioxide (\COtwoE) per passenger~\cite{carboncalculator}\iflater\bcp{Trim
  citation?}\fi---a substantial fraction of the {\em total} yearly emissions
for an average resident of the US (16.5 tons) or Europe (7
tons)~\cite{emissions}.  \iflater\bcp{For comparison, ...  (put in some
  figures about the footprints of a few other things that people might think
  of reducing---driving, eating meat, upgrading their fridge...)}\fi{}
Moreover, these emissions have no near-term 
technological fix, since jet fuel is difficult to replace with
renewable energy sources~\cite{elec-air}. 
% \iflater\bcp{Trimmed this sentence: Reducing the \COtwoE{} footprint
% of conferences will require a reduction in the footprint of the attendees'
% air travel.}\fi

We formed the ACM SIGPLAN Climate Committee with this situation in
mind. We made presentations at SIGPLAN events, had discussions with
ACM and SIG leaders, and ultimately put together a report of ideas for
reducing the \COtwoE{} costs of conference
activities~\cite{ClimateCommitteReport}. The report considers easy 
but incremental changes like making program committee meetings virtual
instead of physical and enabling virtual participation via live-streaming,
all the way to complete re-imagining of conferences and their role in
science (holding conferences completely in virtual spaces, replacing
conferences with journals, etc.), with some intriguing hybrid forms in the
middle (regional conferences, mega-conferences, conferences split into
two locations on different continents, etc.).

\SECTION{The Challenge of CO2 Reduction}

Those we spoke to were almost universally supportive of reducing
\COtwoE{} emissions due to conferences, but very few were eager
to change the way we organize them. This
hesitation makes sense. Science
is a fundamentally social process, and the conference system
accelerates research through high-bandwidth interaction, direct
dissemination of results, network building, and serendipitous
cross-fertilization. How to weigh the benefit to the environment of a
particular \COtwoE-reducing change, such as on-line PC meetings,
against the harm the change could cause to the scientific process,
such as the lost opportunity to build networks and germinate
synergistic ideas?

Scientific value and environmental harm are apples and oranges, so
they are hard to compare. They are also very hard to measure. While we
can reasonably estimate the \COtwoE{} saved by a particular change, it
is harder to guess the impact of the example it sets. This impact is
important. Emissions from air travel to conferences are the lion's
share of ACM's \COtwoE{} footprint, but air travel only contributes 
2\% of current global emissions~\cite{emissions}. This small fraction is not an excuse
for inaction. Just because any individual's or
organization's share of global emissions is miniscule does not mean it
doesn't matter. We must all contribute to the global 40\% reduction
needed by 2030 in a way that makes the most difference, and we must set an example for
others to do the same. It is especially important for scientists---who
as a community have been raising the alarm about climate change---to
not appear to be declaring themselves exempt from cutting back. But
they must do so in a smart way, or else risk disproportionately
harming the progress that science can bring.

@@

Fortunately, there is a well-established mechanism for comparing apples and
oranges---indeed, for comparing almost anything!  Economists call it ``the
magic of markets.''  By mapping choices and preferences in disparate domains
onto a common scale (money), economic markets can be viewed as massively
distributed computations that optimize costs and benefits across vastly
different activities.\bcp{Give a simple, concrete example.}  For this
reason, many policy experts advocate addressing climate change through some
form of carbon tax, which places a monetary cost on greenhouse gas
emissions, sending a clear market signal that changes aggregate
behavior.\bcp{See
  https://kleinmanenergy.upenn.edu/policy-digests/why-carbon-pricing-falls-short
  for lots of good phrasing and citations.}  Exactly how the price of carbon
should be set, and what should be done with the raised funds, is a
challenging policy question, but it has a clear goal: It 
should be chosen so that the world arrives at zero emissions within the
timeframe that climate scientists tell us is safe.

While a global \COtwoE{} market is the ultimate goal, we do not need
to get there in one fell swoop.
Some countries\bcp{e.g.?}, US states (e.g., California), universities
(e.g., UMD and UCLA), and companies (\mwh{Jane Street?}) have
begun to adopt internal carbon pricing policies and markets, to push
their own behavior closer to a sustainable equilibrium. We feel that ACM
should do the same thing. Here's how it might start.

\SECTION{A Carbon Market within ACM}

The main idea is to for ACM to enforce a carbon tax on conference
budgets. At present, ACM charges an ``allocation'' percentage of
12\% on total revenue. It can also impose a tax on
the \COtwoE{} emissions induced by the conference. This tax has two
benefits. First, it creates an incentive for SIGs and conference organizers to
lower total costs by reducing emissions, so they can pay less
tax. They could do this in many ways, such as the ones outlined in our 
report. Second, ACM can use the extra revenue for ``good works.'' Such
good works could directly offset the carbon that paid for them,
or they could provide some other benefit, such as defraying open access
publishing costs.

Getting this idea rolling requires three things: (1) processes to
measure \COtwoE{} used by a conference, for assessing the tax; (2) setting
a tax to the conference budget for the \COtwoE{} it consumed; and (3)
allocating and disbursing collected funds to good works.

For the first, we can start by estimating \COtwoE{} emissions due to
conference travel by analyzing the conference's registration data. We
worked the ACM SGB to develop a \COtwoE{} calculator that takes such
registration data and estimates the \COtwoE{} cost borne by each
individual's travel to the conference. It could become part of
standard practice to run this calculator when closing out the
conference, to assess the tax. \mwh{other \COtwoE{} costs?}
%
Moreover, intrepid SIG leaders and conference organizers can analyze
the gathered data to look for opportunities for reduction. For
example, we used the calculator to compute the per-capita and total
\COtwoE{} costs of several SIGPLAN conferences over the last few years.
\mwh{Discuss.}
\bcp{Put the discussion of Figure 1 (and an accompanying figure
showing distance traveled to several instances of the same conference) here;
this illustrates the difficulty and the need for careful data gathering and
analysis.}

@@

How should we set the price of a conference's \COtwoE? There are several
possibilities. One idea is simply to adopt the price from an existing
market, such as California's; between 2012 and 2018, the price of a
ton of \COtwoE{} fluctuated from about \$12 to \$23~\cite{CACO2}. Another idea is to
match the price to the equivalent in \emph{carbon
  offsets}~\cite{CarbonOFfsetReport}. Offset providers use raised funds
to support an activity (e.g., installing methane capture devices on
landfill sites or buying fuel-efficient stoves to replace open-fire
cooking in poor communities) that is {\em certified} (by yet another
organization) to permanently avoid emitting an amount of greenhouse
gases specified. Current prices vary widely around an average of around
\$15 per ton, depending on the vendor, the nature of the offset
project, the region of the world where it will be implemented, etc.

However we estimate the price, we could imagine setting it a little low at the
outset, perhaps based on analysis of existing conference \COtwoE{} usage, and then raising it
over time as organizers start to adapt to it. On the other hand, the true
``social cost'' of a ton of carbon is variously estimated to be much
higher than the above---anywhere from \$40 to \$400 per
ton~\bcp{citation}---so current pricing may already be on the low
side.

Lastly, having collected the carbon tax from the conference budget,
what should ACM do with it? One obvious thing to do is to buy \COtwoE{}
offsets.\footnote{Some ACM conferences have already started to
  experiment with voluntary carbon offset fees at registration time,
  but with little luck.}
%
Carbon offsets are not a long-term solution to global warming, and the
details matter, but at this point we believe there are options that do
far more good than harm.
%
Many organizations (including companies such as Google, Dell, Microsoft,
General Motors, Delta Airlines, Lyft, and Expedia, numerous universities,
and a number of academic societies) are already buying carbon offsets.
%
One superficial impediment is that most organizations do not currently have
accounting procedures in place to deal with them, and the piecemeal
political and bureaucratic work of designing and implementing a separate
policy for each organization can consume significant energy. \mwh{Not
  sure the previous is the right thing to say? Basically: Need to work
  out the legality. Cite offset report?}

Another thing is to do \mwh{finish off. Here, talk about other good
  works, like open access fees, developing technology for CO2e
  reduction of hosting conferences, perhaps pay for remote presence,
  ...}

\SECTION{Conclusion}
